---
title: "introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(flipr)
```

## Usage ##

### Creating a transition list with LipidCreator



### Measuring a standard sample on an MS platform

### Converting the raw MS data

We use msConvert to convert the raw MS data for different vendor platforms into mzML.


### Extracting fragment ion traces from the MS file 

Transition extractor 

### Anatomy of the feature table format (*-fip.tsv)

The feature table `*-fip.tsv` is created by the transition extraction step and serves as the main data input for flipR- It holds one m/z feature per row, with the following columns, specifying its provenance, parameters and information that is used by downstream steps to maintain a mapping between input transition list and output parameter file for LipidCreator- The following columns are reported in the output file (some correspond to *.mzML elements / attributes, some come from LipidCreator / Skyline):

- instrument
- localDateTimeCreated
- origin
- scanNumber
- polarity
- basePeakMz
- basePeakIntensity
- totalIonCurrent
- id
- scanDefinition
- msLevel
- isolationWindowTargetMz[0]
- isolationWindowLowerOffset[0]
- isolationWindowUpperOffset[0]
- precursorActivationType
- precursorCollisionEnergy
- precursorCollisionEnergyUnit
- ionInjectionTime[0]
- isolationMzMin[0]
- isolationMzMax[0]
- precursorCharge[0]
- precursorMz[0]
- msFunction
- retentionTime
- spectrumType
- rawTic
- group
- foundMass
- foundMassRange[ppm]
- foundMassLowerBound
- foundMassUpperBound
- foundMassError[ppm]
- foundIntensity
- scanRelativeIntensity
- calculatedMass
- species
- precursorAdduct
- fragment
- adduct

### Fitting a model for multiple fragment ion traces over a CE sequence

```R
  library("flipr")
  xics_file <- system.file("extdata", "12-HETE-d8-_M-H_1--qex_fip.tsv", package = "flipr", mustWork = TRUE)
  fits <- flip(projectDir = dirname(xics_file), plotFormat = "pdf", filePattern = basename(xics_file), dataPlots = TRUE)
```

### LipidCreator Parameter Table (After model training and selection)
The comma-separated lipid creator parameter file contains collision energy calculation parameters for each lipid class, as reported and concatenated by flipR- There are as many rows for each fragment, as there are parameters- The following columns are required in this file: 
- instrument the PSI-MS CV term id identifying the instrument, e.g- MS:1002523 for Thermo Scientific Q Exactive HF.
- class	the lipid class, e.g- 10-HDoHE, needs to be double quoted, when the name contains a comma.
- adduct the precursor adduct for this lipid class, e.g- [M-H]1-.
- fragment the fragment identifier- If no common name is available, use e.g- "m/z 121.0658"- The precursor must be reported as "precursor".
- ParKey the model parameter name, currently one of "model", "meanlog", "sdlog", "scale", and "shift".
- ParValue the model parameter values, currently, for "model" only "dlnormPar" is recognized- Other parameters are expected to be reported as double numbers with a "." as the decimal separator.